### 2019-06-09 2-Chapters/4-Chapter/IEEE_WCNC_2019__DemoICT.git/main.tex
Done, complete proof-reading of Chapter:4

In this demo, we can generate scenarios with different parameters (number of channels, interfering traffic, etc) in order to evaluate the performance of learning in various settings.
Moreover, we compare the performance of learning strategies with that of the random uniform access to channels, which is the current state-of-the-art of commercial LPWAN solutions \cite{Raza17}.
%
This allows to check that in case of uniform traffic, when there is nothing to learn, the intelligent devices at least do not reduce their successful communication rate in comparison to the naive devices.
This also shows that in case of non-uniform stationary traffic, MAB learning algorithms indeed help to increase the global efficiency of the network by improving the success rate of the intelligent devices.
%
The benefits are twice and of primary importance for IoT networks:
the proposed approach can mitigate RF collisions,
and enhance intelligent device battery lifetime if they do retransmissions.

@@

\paragraph{Details about our implementation.}

We used the GNU Radio Companion software (GRC, version $3.7$ in $2017$),
and a laptop runs a GRC design to configure and control each USRP platform.
As such, one laptop can run in parallel the control program of any number of boards\footnote{~Even if in practice, maximum efficiency is kept as long as there is not more than one GRC design by CPU core.}.
%
The GNU Radio software provides the framework and tools to build and run software radio or just general signal-processing applications.
GNU Radio applications are flow-graphs: a series of signal processing blocks connected together to describe a data flow.
For maximum efficiency, we wrote all of our blocks in \texttt{C++}.
These flow-graphs can be written in either \texttt{C++} or the Python programming language. The GNU Radio infrastructure is written entirely in \texttt{C++}, and many of the user tools are written in Python.
GNU Radio Companion is a graphical user interface (UI) used to develop GNU Radio applications:
GRC is effectively a Python code-generation tool.
When a flow-graph is compiled in GRC, a Python code is produced, which can be executed to connect to the USRP,
create the desired GUI windows and widgets, and create and connect the blocks in the flow-graph.

@@

% ----------------------------------------------------------------------
\subsection{Experimental results}
\label{sub:42:results}
% ----------------------------------------------------------------------
We compare in this PoC the two algorithms described in Section~\ref{sec:2:famousMABalgorithms} (\UCB{} and Thompson sampling) against a uniform access algorithm, that uniformly selects its channel at random.
For one dynamic device, three algorithms are compared by their mean successful communication rates, on a horizon of $T=2000$ communication slots, and were using three algorithms: uniform random access (in \textcolor{cyan}{cyan}), Thompson Sampling (in \textcolor{green}{green}) and \UCB{} (in \textcolor{red}{red}).

@@

In Figure~\ref{fig:42:plot_datafile_append_Uniform_vs_UCB_vs_TS} above, we show the results averaged on $10$ repetitions using the same conditions.
%
Each experiment takes about half a day,
as we make devices generate one message every $5$ seconds, in order to artificially speed up the process and with no loss of generality (as we are using real hardware).
Learning can be useful only when there is a large enough difference between ``good'' and ``bad'' channels,
Each device was learning to access $4$ different non-overlapping channels, that we chose to have occupancy rates of $(\mu_i)_i = [15\%, 10\%, 2\%, 1\%]$.
The occupancy rate of a channel, which denotes the mean occupancy, is implemented using the traffic generator, to emulate the presence of $S_i$ static devices with emission probability $p$ (that is, $\mu_i = S_i \times p$ here).
% They did not overlap each other, as
When facing the same stationary background traffic, we see that the learning devices are both very quickly more efficient than the naive uniform device.
We obtain an improvement in terms of successful communication rate from $40\%$ to about $60\%$ in only $100$ communications (about $16\;\mathrm{min}$), and up-to $80\%$ in only $400$ communications.
%
In stationary environments, both the TS and \UCB{} algorithms are very efficient and converge quickly, resulting in a very strong decrease in collisions and failed communication slots. \UCB{} is faster to learn but eventually TS gives a (slightly) better average performance.

>>>>

We can generate scenarios with different parameters (number of channels, interfering traffic, etc) in order to evaluate the performance of learning in various settings.
Moreover, we compare the performance of learning strategies with that of the random uniform access to channels, which is the current state-of-the-art of commercial LPWAN solutions \cite{Raza17}.
%
This allows to check that in case of uniform traffic, when there is nothing to learn, the intelligent devices at least do not reduce their successful communication rate in comparison to the naive devices.
This also shows that in case of non-uniform stationary traffic, MAB learning algorithms indeed help to increase the global efficiency of the network by improving the success rate of the intelligent devices.
%
The benefits are twice and of primary importance for IoT networks:
the proposed approach can mitigate RF collisions,
and enhance intelligent device battery lifetime if they do retransmissions.

@@

\paragraph{Details about our implementation.}
%
We used the GNU Radio Companion software (GRC, version $3.7$ in $2017$),
and a laptop runs a GRC design to configure and control each USRP platform.
As such, one laptop can run in parallel the control program of any number of boards\footnote{~Even if in practice, maximum efficiency is kept as long as there is not more than one GRC design by CPU core.}.
%
The GNU Radio software provides the framework and tools to build and run software radio or just general signal-processing applications.
GNU Radio applications are flow-graphs: a series of signal processing blocks connected together to describe a data flow.
For maximum efficiency, we wrote all of our blocks in \texttt{C++}.
These flow-graphs can be written in either \texttt{C++} or the Python programming language. The GNU Radio infrastructure is written entirely in \texttt{C++}, and many of the user tools are written in Python.
GNU Radio Companion is a graphical user interface (UI) used to develop GNU Radio applications:
GRC is effectively a Python code-generation tool.
When a flow-graph is compiled in GRC, a Python code is produced, which can be executed to connect to the USRP,
create the desired GUI windows and widgets, and create and connect the blocks in the flow-graph.

@@

% ----------------------------------------------------------------------
\subsection{Experimental results}
\label{sub:42:results}
% ----------------------------------------------------------------------
We compare in this PoC the two algorithms described in Section~\ref{sec:2:famousMABalgorithms} (\UCB{} and Thompson sampling) against a uniform access algorithm, that uniformly selects its channel at random.
For one dynamic device, three algorithms are compared by their mean successful communication rates, on a horizon of $T=2000$ communication slots, and were using three algorithms: uniform random access (in \textcolor{cyan}{cyan}), Thompson Sampling (``\textcolor{green}{TS}'', in \textcolor{green}{green}) and \UCB{} (in \textcolor{red}{red}).

@@

In Figure~\ref{fig:42:plot_datafile_append_Uniform_vs_UCB_vs_TS} below, we show the results averaged on $10$ repetitions using the same conditions.
%
Each experiment takes about half a day,
as we make devices generate one message every $5$ seconds, in order to artificially speed up the process and with no loss of generality (as we are using real hardware).
Learning can be useful only when there is a large enough difference between ``good'' and ``bad'' channels,
Each device was learning to access $4$ different non-overlapping channels, that we chose to have occupancy rates of $(\mu_i)_i = [15\%, 10\%, 2\%, 1\%]$.
The occupancy rate of a channel, which denotes the mean occupancy, is implemented using the traffic generator, to emulate the presence of $S_i$ static devices with emission probability $p$ (that is, $\mu_i = S_i \times p$ here).
% They did not overlap each other, as
When facing the same stationary background traffic, we see that the learning devices are both very quickly more efficient than the naive uniform device.
We obtain an improvement in terms of successful communication rate from $40\%$ to about $60\%$ in only $100$ communications (about $16\;\mathrm{min}$), and up-to $80\%$ in only $400$ communications.
%
In stationary environments, both the TS and \UCB{} algorithms are very efficient and converge quickly, resulting in a very strong decrease in collisions and failed communication slots. \UCB{} is faster to learn but eventually TS gives a (slightly) better average performance.

### 2019-06-09 2-Chapters/4-Chapter/IEEE_WCNC__2019__Paper__BMBM.git/IEEE_WCNC__2019__Paper__BMBM.tex
Done, complete proof-reading of Chapter:4

In this section, we extend the model presented before, to take into account the possibility for retransmissions of a message after a collision.
This extension is of major importance: most protocols for real-world IoT networks can use retransmissions, it is for instance the case of the LoRaWAN standard.
As before, we propose and evaluate different learning strategies based on MAB algorithms.
However, the price to be paid is a shorter battery lifetime for IoT devices. So this approach would be used only for devices with strong delivery constraints (\eg, for health-care applications), or that can refuel their energy over time (\eg, for robotic applications).

@@

\textbf{Outline.}
%
The rest of the section is organized as follows.
First the system model is introduced in Section~\ref{sub:43:model},
and our motivations are exposed in Section~\ref{sub:43:motivations}.
The proposed \UCB-based heuristics are presented in Section~\ref{sub:43:heuristics}, while the corresponding numerical results are shown in Section~\ref{sub:43:numExp}.
Finally, some conclusions are drawn later in Section~\ref{sub:43:conclusion}.

@@

% % \begin{small} % XXX remove if needed
% \begin{figure}[h!]
% 	\centering
%     \begin{framed}
% 	\begin{algorithm}[H]
% 		% \begin{small} % XXX remove if needed
% 		\For(){$t = 1, \dots, T$}{
% 			Compute $\forall k, U_k(t) = \widehat{\mu_k}(t) + \sqrt{\alpha \log(t) / N_k(t)}$\;
% 			Transmit in channel $A(t) \sim \cU(\argmax_k U_k(t))$\;
% 			Reward $r_{A(t)}(t) = 1$, if \emph{Ack} is received, else $0$\;
% 		}
% 		\caption{The \UCB{} algorithm for channel selection (base building block).}
% 		\label{algo:43:UCB}
% 		% \end{small} % XXX remove if needed
% 	\end{algorithm}
% 	\end{framed}
% \end{figure}
% % \end{small} % XXX remove if needed

% ----------------------------------------------------------------------
\subsection{The first heuristics: \UCB{} unaware of retransmissions}
\label{sub:43:UCBnaive}
% ----------------------------------------------------------------------

@@

This first heuristic we propose is unaware of retransmission: the same channel is used for retransmissions.
The \UCB{} algorithm is implemented independently by each device, we denote it ``first-stage'' \UCB, and we present it in Algorithm~\ref{algo:43:UCB}.
It is the same algorithm as the Algorithm~\ref{algo:2:indexPolicy} in Section~\ref{sub:2:IndexPolicies} (using the same \UCB{} indexes as we defined them in equation \eqref{eq:UCB_index}),
but we write it again to make clear the difference with first transmission and retransmission of messages.
%
Note that a device using this first approach is only able to select a channel for the first transmission (using \UCB), and it uses the same channel for all the corresponding retransmissions of a packet (if retransmissions happen).

% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
    \begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		\For(){$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Compute $\forall k, U_k(t) = \widehat{\mu_k}(t) + \sqrt{\alpha \log(t) / N_k(t)}$\;
				Transmit in channel $A(t) \sim \cU(\argmax_k U_k(t))$\;
				Reward $r(t) = 1$, if \emph{Ack} is received, else $0$\;
			}
			\Else(\tcp*[f]{Retransmit in same channel}){
				j $\leftarrow$ last channel selected by first-stage \UCB\;
				Transmit in channel $A(t) = j$\;
			}
		}
		\caption{First-stage \UCB{} and retransmission in same channel.}
		\label{algo:43:UCB}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

More formally, for one device, let $N_k(t)$ be the number of times the channel $k$ (for $k\in [K]$) was selected up-to time $t-1$, for $t\geq 0$
for any $t\in\mathbb{N}$,
$N_k(t) = \sum_{\tau=0}^{t-1} \mathbbm{1}(A(\tau) = k)$.
The empirical mean estimator $\widehat{\mu_k}(t)$ of channel $k$ is defined as the mean reward obtained up-to time $t-1$,
$\widehat{\mu_k}(t) = \frac{1}{N_k(t)} \sum_{\tau=0}^{t-1} r(\tau) \mathbbm{1}(A(\tau) = k)$,
where $r(t)=Y_{k,t}$ is the reward obtained after transmission in channel $k$ at time $t$.
The upper confidence bound in each channel $k$ is defined as
$U_k(t) = \widehat{\mu_k}(t) + \sqrt{\alpha \log(t) / N_k(t)}$.
Finally, the transmission channel at time step $t$
is the one maximizing this \UCB{} index $U_k(t)$,
as it is the one expected to be the best one at the current time step $t$,
$A(t) = \arg\max_k U_k(t)$.

\begin{leftbar}[warningbar]  % XXX leftbar warningbar, comment if needed
	\textbf{\textcolor{red}{Small warning about notations: local (device) time vs global (gateway) time}:}
	%
	In all the algorithms and notations presented in this Section, the time steps $t$ do not refer to the global time steps (as seen from the gateway), but rather the current number of uplink transmissions (or retransmissions) that was carried out by the device.
	Each device wakes up whenever it has to send some data, following its random emission process (here, we propose a Bernoulli process of small probability, \eg, $p=10^{-3}$), and when it woke up, it sends its packet, and will send again for at most $M$ times (\eg, $M=10$) until it receives an \Ack.
	%
	In other words, the time steps $t$ in the following algorithms denote the number of time the device sent an uplink packet, and these uplink (re)transmissions can happen at different (real-world) times, more or less spread out in time, it does not matter to the learning part of this model and work.
\end{leftbar}  % XXX leftbar warningbar, comment if needed

@@

% --- --- --- --- --- --- --- --- --- ---
\paragraph{Uniform random retransmission ``\textcolor{cyan}{(Random)''}.}\label{sub:43:UCBthenRandom}

In this first proposal, the device uses a random channel selection, following a uniform distribution (on $[K]$).
It is described below in Algorithm~\ref{algo:43:UCBthenRandom}.
More precisely, the first-stage \UCB{} use rewards built from the acknowledgments that the device received or not for its first-stage transmission, but do not use any feedback about any of the retransmissions of any message.

% \vspace*{-3pt}
\begin{figure}[h!]
	\centering
    \begin{framed}
	\begin{algorithm}[H]
	% \begin{small} % XXX remove if needed
	\For(){$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Use first-stage \UCB{} as in Algorithm~\ref{algo:43:UCB}.
			}
			\Else(\tcp*[f]{Random retransmission}){
				Transmit in channel $A(t) \sim \mathcal{U}(1,\ldots,K)$\;
			}
		}
		\caption{Uniform random retransmission.}    % A naive
		\label{algo:43:UCBthenRandom}
	% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}

% --- --- --- --- --- --- --- --- --- ---
\paragraph{\UCB{} for retransmission (``\textcolor{blue}{Only \UCB{}}'').}\label{sub:43:TwoUCB}

Instead of applying a random channel selection for retransmission,
another heuristic is to use a second \UCB{} algorithm in the second stage.
In other words, we expect that this algorithm is able to learn the best channel to retransmit a packet.
It is described in Algorithm~\ref{algo:43:TwoUCB}, and it is still a practical approach, since the storage requirements and time complexity remains linear w.r.t. the number of channels $K$ (\ie, $\mathcal{O}(K)$).
%
Note that, we use the subscript $({}^r)$ to denote the variables
$\widehat{\mu^r}(t)$, $B^r_k(t)$ and $U^r_k(t)$,
related to the \UCB{} algorithm employed for the retransmission.

% \vspace*{-3pt}
% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
    \begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		\For(){$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Use first-stage \UCB{} as in Algorithm~\ref{algo:43:UCB}.
			}
			\Else(\tcp*[f]{Packet retransmission with $\mathrm{UCB}^r$}){
				Compute $\forall k, U^r_k(t) = \widehat{\mu^r_k}(t) + \sqrt{\alpha \log(t) / N_k^r(t)}$\;
				Transmit in channel $C^r(t) \sim \cU(\argmax_k U^r_k(t))$\;
				Reward $r^r(t) = 1$, if \emph{Ack} is received, else $0$\;
			}
		}
		\caption{\UCB{} for retransmission.}    % A naive
		\label{algo:43:TwoUCB}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

% --- --- --- --- --- --- --- --- --- ---
\paragraph{$K$ different {\UCB}s for retransmission (``\textcolor{green}{$K$ \UCB}'')}\label{sub:43:UCBthenKp1}

Another heuristic is to not use the same algorithm no matter where the collision occurred, but to use $K$ different second-stage \UCB{} algorithms.
It means that after a failed first transmission in channel $j$, the device relies on the $j$-th algorithm to decide its retransmission.
The corresponding algorithm is depicted in Algorithm~\ref{algo:43:UCBthenKp1}.
Each of these algorithms are denoted using the subscript $({}^{j})$, for $j\in[K]$.

% \vspace*{-3pt}
% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
\begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		% \KwData{$\forall k,j\in[\![1;K]\!]$, $N_k^j(t)$, $\widehat{\mu_k}^j(t)$, $B_k^j(t)$ and $U_k^j(t)$}
		\For(\tcp*[f]{At every time step})
		{$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Use first-stage \UCB{} as in Algorithm~\ref{algo:43:UCB}.
			}
			\Else(\tcp*[f]{Packet retransmission with $\mathrm{UCB}^j$}){ % Retr. after trying channel $j$
				j $\leftarrow$ last channel selected by first-stage \UCB\;
				Compute $\forall k, U_k^j(t) = \widehat{\mu_k}^j(t) + \sqrt{\alpha \log(t) / N_k^j(t)}$\;
				Transmit in channel $A^j(t) \sim \cU(\argmax_k U^j_k(t))$\;
				Reward $r^j_{A^j(t)}(t) = 1$ if \emph{Ack} is received, else $0$\;
			}
		}
		\caption{$K$ different {\UCB}s for retransmission.}
		\label{algo:43:UCBthenKp1}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

@@

% --- --- --- --- --- --- --- --- --- ---
\paragraph{Delayed \UCB{} for retransmission (``\textcolor{red}{Delayed \UCB}'').}\label{sub:43:UCBwithDelay}

This last heuristic is a composite of
the random retransmission (Algorithm~\ref{algo:43:UCBthenRandom})
and the \UCB{} retransmission (Algorithm~\ref{algo:43:TwoUCB}) approaches.
Instead of starting the second stage \UCB{} directly from the first retransmission, we introduce a fixed delay $\Delta\in\mathbb{N}$, $\Delta \geq 1$,
and start to rely on the second stage \UCB{} after $\Delta$ transmissions.
The selection for the first steps is handled with the random retransmission.

@@

% \vspace*{-3pt}
% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
	\begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		\For(\tcp*[f]{At every time step})
		{$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Use first-stage \UCB{} as in Algorithm~\ref{algo:43:UCB}.
			}
			\uElseIf(\tcp*[f]{Random retransmission}){$t \leq \Delta$}{
				Transmit randomly in a channel $A(t) \sim \mathcal{U}(1,\ldots,K)$.
			}
			\Else(\tcp*[f]{Packet retransmission with delayed $\mathrm{UCB}^d$}){
				Compute $\forall k, U^d_k(t) = \widehat{\mu^d_k}(t) + \sqrt{\alpha \log(t) / N_k^d(t)}$\;
				Transmit in channel $A^d(t) \sim \cU(\argmax_k U^d_k(t))$\;
				Reward $r^d_{A^d(t)}(t) = 1$ if \emph{Ack} is received, else $0$\;
			}
		}
		\caption{Delayed \UCB{} for retransmission.}
		\label{algo:43:UCBwithDelay}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

\paragraph{Other ideas that we did not explore.}
%
Instead of considering a first-stage \UCB{} followed by a random channel retransmission, we could have considered a random first-stage channel retransmission followed by a second-stage \UCB.
This was not considered in our study \cite{Bonnefoi2019WCNC}, and we preferred to not add it to the experiments presented below, for three reasons: to avoid clutter in the plots, to win some time as these experiments had to run for quite a long time, but mainly because the first-stage channel selection is most important one as we illustrate by the large difference in terms of performance between the uniform channel selection and the ``Only UCB'' heuristic, below in Figures~\ref{fig:43:mainExperiment1} and \ref{fig:43:mainExperiment2}.

>>>>

In this section, we extend the previous model to take into account the possibility for retransmissions of a message after a collision.
This is of major importance as most protocols for real-world IoT networks can use retransmissions, it is for instance the case of the LoRaWAN standard.
As before, we propose and evaluate different learning strategies based on MAB algorithms.
However, the price to be paid is a shorter battery lifetime for IoT devices. So this approach would be used only for devices with strong delivery constraints (\eg, for health-care applications), or that can refuel their energy over time (\eg, for robotic applications).

@@

% \textbf{Outline.}
% %
% The rest of the section is organized as follows.
% First the system model is introduced in Section~\ref{sub:43:model},
% and our motivations are exposed in Section~\ref{sub:43:motivations}.
% The proposed \UCB-based heuristics are presented in Section~\ref{sub:43:heuristics}, while the corresponding numerical results are shown in Section~\ref{sub:43:numExp}.
% Finally, some conclusions are drawn later in Section~\ref{sub:43:conclusion}.

@@

% % \begin{small} % XXX remove if needed
% \begin{figure}[h!]
% 	\centering
%     \begin{framed}
% 	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
% 		\For(){$t = 1, \dots, T$}{
% 			Compute $\forall k, U_k(t) = \widehat{\mu_k}(t) + \sqrt{\alpha \log(t) / N_k(t)}$\;
% 			Transmit in channel $A(t) \sim \cU(\argmax_k U_k(t))$\;
% 			Reward $r_{A(t)}(t) = 1$, if \emph{Ack} is received, else $0$\;
% 		}
% 		\caption{The \UCB{} algorithm for channel selection (base building block).}
% 		\label{algo:43:UCB}
		% \end{small} % XXX remove if needed
% 	\end{algorithm}
% 	\end{framed}
% \end{figure}
% % \end{small} % XXX remove if needed

% ----------------------------------------------------------------------
\subsection{The first heuristic: \UCB{} unaware of retransmissions}
\label{sub:43:UCBnaive}
% ----------------------------------------------------------------------

@@

This first heuristic we propose is unaware of retransmission: the same channel is used for retransmissions.
The \UCB{} algorithm is implemented independently by each device, we denote it ``first-stage'' \UCB, and we present it in Algorithm~\ref{algo:43:UCB}.
It is the same algorithm as the Algorithm~\ref{algo:2:indexPolicy} in Section~\ref{sub:2:IndexPolicies} (using the same \UCB{} indexes as we defined them in equation \eqref{eq:UCB_index}),
but we write it again to make clear the difference with first transmission and retransmission of messages.
%
Note that a device using this first approach is only able to select a channel for the first transmission (using \UCB, line 3-5), and then it uses the same channel for all the corresponding retransmissions of a packet (if retransmissions happen, line 7-8).

% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
    \begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		\For(){$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Compute $\forall k, U_k(t) = \widehat{\mu_k}(t) + \sqrt{\alpha \log(t) / N_k(t)}$\;
				Transmit in channel $A(t) \sim \cU(\argmax_k U_k(t))$\;
				Reward $r(t) = 1$, if \emph{Ack} is received, else $0$\;
			}
			\Else(\tcp*[f]{Retransmit in same channel}){
				j $\leftarrow$ last channel selected by first-stage \UCB\;
				Transmit in channel $A(t) = j$\;
			}
		}
		\caption[First-stage \UCB{} and retransmission in same channel.]{First-stage \UCB{} and retransmission in same channel (``\textcolor{blue}{Only \UCB{}}'').}
		\label{algo:43:UCB}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

More formally, for one device, let $N_k(t)$ be the number of times the channel $k$ (for $k\in [K]$) was selected up-to time $t-1$, for $t\geq1$
for any $t\in\mathbb{N}$,
$N_k(t) = \sum_{\tau=1}^{t-1} \mathbbm{1}(A(\tau) = k)$.
The empirical mean estimator $\widehat{\mu_k}(t)$ of channel $k$ is defined as $\widehat{\mu_k}(t) = \frac{1}{N_k(t)} \sum_{\tau=1}^{t-1} r(\tau) \mathbbm{1}(A(\tau) = k)$,
where $r(t)=Y_{k,t}$ is the reward obtained after transmission in channel $k$ at time $t$.
The upper confidence bound in each channel $k$ is defined as
$U_k(t) = \widehat{\mu_k}(t) + \sqrt{\alpha \log(t) / N_k(t)}$.
Finally, the transmission channel at time step $t$
is $A(t) \sim \cU(\argmax_k U_k(t))$.
% the one maximizing this \UCB{} index $U_k(t)$,
% as it is the one expected to be the best one at the current time step $t$,

\begin{leftbar}[warningbar]  % XXX leftbar warningbar, comment if needed
	\textbf{\textcolor{red}{Small warning about notations: local (device) time vs global (gateway) time}:}
	%
	In all the algorithms and notations presented in this Section, the time steps $t$ do not refer to the \emph{global} time steps (as seen from the gateway), but rather the current number of uplink transmissions (or retransmissions) that was carried out by the device.
	Each device wakes up whenever it has to send some data, following its random emission process (here, we propose a Bernoulli process of small probability, \eg, $p=10^{-3}$), and when it woke up, it sends its packet, and will send again for at most $M$ times (\eg, $M=10$) until it receives an \Ack.
	%
	In other words, the time steps $t$ in the following algorithms denote the number of time the device sent an uplink packet, and these uplink (re)transmissions can happen at different (real-world) times, more or less spread out in time, it does not matter to the learning part of this model and work.
\end{leftbar}  % XXX leftbar warningbar, comment if needed

@@

% --- --- --- --- --- --- --- --- --- ---
\textbf{Uniform random retransmission ``(\textcolor{cyan}{Random}'').}\label{sub:43:UCBthenRandom}
%
In this first proposal, the device uses a random channel selection, following a uniform distribution (on $[K]$).
It is described below in Algorithm~\ref{algo:43:UCBthenRandom}.
More precisely, the first-stage \UCB{} use rewards built from the acknowledgments that the device received or not for its first-stage transmission, but do not use any feedback about any of the retransmissions of any message.

% \vspace*{-3pt}
\begin{figure}[h!]
	\centering
    \begin{framed}
	\begin{algorithm}[H]
	% \begin{small} % XXX remove if needed
	\For(){$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Use first-stage \UCB{} as in Algorithm~\ref{algo:43:UCB}.
			}
			\Else(\tcp*[f]{Random retransmission}){
				Transmit in channel $A(t) \sim \mathcal{U}(1,\ldots,K)$\;
			}
		}
		\caption[Heuristic: uniform random retransmission.]{Heuristic: uniform random retransmission ``(\textcolor{cyan}{Random}'').}    % A naive
		\label{algo:43:UCBthenRandom}
	% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}

% --- --- --- --- --- --- --- --- --- ---
\textbf{\UCB{} for retransmission (``\textcolor{purple}{\UCB{}}'').}\label{sub:43:TwoUCB}
%
Instead of applying a random channel selection for retransmission,
another heuristic is to use a second \UCB{} algorithm in the second stage.
In other words, we expect that this algorithm is able to learn the best channel to retransmit a packet.
It is described in Algorithm~\ref{algo:43:TwoUCB}, and it is still a practical approach, since the storage requirements and time complexity remains linear w.r.t. the number of channels $K$ (\ie, $\mathcal{O}(K)$).
%
Note that, we use the subscript $({}^r)$ to denote the variables
$\widehat{\mu^r}(t)$, $B^r_k(t)$ and $U^r_k(t)$,
related to the \UCB{} algorithm employed for the retransmission.

% \vspace*{-3pt}
% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
    \begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		\For(){$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Use first-stage \UCB{} as in Algorithm~\ref{algo:43:UCB}.
			}
			\Else(\tcp*[f]{Packet retransmission with $\mathrm{UCB}^r$}){
				Compute $\forall k, U^r_k(t) = \widehat{\mu^r_k}(t) + \sqrt{\alpha \log(t) / N_k^r(t)}$\;
				Transmit in channel $C^r(t) \sim \cU(\argmax_k U^r_k(t))$\;
				Reward $r^r(t) = 1$, if \emph{Ack} is received, else $0$\;
			}
		}
		\caption[Heuristic: \UCB{} for retransmission.]{Heuristic: \UCB{} for retransmission (``\textcolor{purple}{\UCB{}}'').}    % A naive
		\label{algo:43:TwoUCB}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

% --- --- --- --- --- --- --- --- --- ---
\textbf{$K$ different {\UCB}s for retransmission (``\textcolor{green}{$K$ \UCB}'')}\label{sub:43:UCBthenKp1}
%
Another heuristic is to not use the same algorithm no matter where the collision occurred, but to use $K$ different second-stage \UCB{} algorithms.
It means that after a failed first transmission in channel $j$, the device relies on the $j$-th algorithm to decide its retransmission.
The corresponding algorithm is depicted in Algorithm~\ref{algo:43:UCBthenKp1}.
Each of these algorithms are denoted using the subscript $({}^{j})$, for $j\in[K]$.

% \vspace*{-3pt}
% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
\begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		% \KwData{$\forall k,j\in[\![1;K]\!]$, $N_k^j(t)$, $\widehat{\mu_k}^j(t)$, $B_k^j(t)$ and $U_k^j(t)$}
		\For(\tcp*[f]{At every time step})
		{$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Use first-stage \UCB{} as in Algorithm~\ref{algo:43:UCB}.
			}
			\Else(\tcp*[f]{Packet retransmission with $\mathrm{UCB}^j$}){ % Retr. after trying channel $j$
				j $\leftarrow$ last channel selected by first-stage \UCB\;
				Compute $\forall k, U_k^j(t) = \widehat{\mu_k}^j(t) + \sqrt{\alpha \log(t) / N_k^j(t)}$\;
				Transmit in channel $A^j(t) \sim \cU(\argmax_k U^j_k(t))$\;
				Reward $r^j_{A^j(t)}(t) = 1$ if \emph{Ack} is received, else $0$\;
			}
		}
		\caption[Heuristic: $K$ different {\UCB}s for retransmission.]{Heuristic: $K$ different {\UCB}s for retransmission (``\textcolor{green}{$K$ \UCB}'').}
		\label{algo:43:UCBthenKp1}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

@@

% --- --- --- --- --- --- --- --- --- ---
\textbf{Delayed \UCB{} for retransmission (``\textcolor{red}{Delayed \UCB}'').}\label{sub:43:UCBwithDelay}
%
This last heuristic is a composite of
the random retransmission (Algorithm~\ref{algo:43:UCBthenRandom})
and the \UCB{} retransmission (Algorithm~\ref{algo:43:TwoUCB}) approaches.
Instead of starting the second stage \UCB{} directly from the first retransmission, we introduce a fixed delay $\Delta\in\mathbb{N}$, $\Delta \geq 1$,
and start to rely on the second stage \UCB{} after $\Delta$ transmissions.
The selection for the first steps is handled with the random retransmission.

@@

% \vspace*{-3pt}
% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
	\begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		\For(\tcp*[f]{At every time step})
		{$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Use first-stage \UCB{} as in Algorithm~\ref{algo:43:UCB}.
			}
			\uElseIf(\tcp*[f]{Random retransmission}){$t \leq \Delta$}{
				Transmit randomly in a channel $A(t) \sim \mathcal{U}(1,\ldots,K)$.
			}
			\Else(\tcp*[f]{Packet retransmission with delayed $\mathrm{UCB}^d$}){
				Compute $\forall k, U^d_k(t) = \widehat{\mu^d_k}(t) + \sqrt{\alpha \log(t) / N_k^d(t)}$\;
				Transmit in channel $A^d(t) \sim \cU(\argmax_k U^d_k(t))$\;
				Reward $r^d_{A^d(t)}(t) = 1$ if \emph{Ack} is received, else $0$\;
			}
		}
		\caption[Heuristic: delayed \UCB{} for retransmission.]{Heuristic: delayed \UCB{} for retransmission (``\textcolor{red}{Delayed \UCB}'').}
		\label{algo:43:UCBwithDelay}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

\textbf{Other ideas that we did not explore.}
%
Instead of considering a first-stage \UCB{} followed by a random channel retransmission, we could have considered a random first-stage channel retransmission followed by a second-stage \UCB.
This was not considered in our study \cite{Bonnefoi2019WCNC}, and we preferred to not add it to the experiments presented below, for three reasons: to avoid clutter in the plots, to win some time as these experiments had to run for quite a long time, but mainly because the first-stage channel selection is most important one as we illustrate by the large difference in terms of performance between the uniform channel selection and the ``Only UCB'' heuristic, below in Figures~\ref{fig:43:mainExperiment1} and \ref{fig:43:mainExperiment2}.
